---
title: "browse exp of 52"
output: html_document
date: '2023-02-21'
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

library(tidyverse)
library(RColorBrewer)
library(data.table)
library(edgeR)
library(qvalue)
library(Heatplus)
library(gplots)
library(ggrepel)
library(knitr)
library(ggnewscale)

# define some useful funcs
sample_n_of <- function(data, size, ...) {
  dots <- quos(...)
  
  group_ids <- data %>% 
    group_by(!!! dots) %>% 
    group_indices()
  
  sampled_groups <- sample(unique(group_ids), size)
  
  data %>% 
    filter(group_ids %in% sampled_groups)
}


# Set theme
theme_set(
  theme_classic() +
  theme(text=element_text(size=16,  family="Helvetica")))

# I use layer a lot, to rotate long x-axis labels
Rotate_x_labels <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

#test plot
ggplot(mtcars, aes(x=mpg, y=cyl)) +
  geom_point()


```

## Intro

Jingxin's lab has synthesized (or maybe contracted out synthesis of) hundred(s) branaplam/risdiplam derivatives. They screened them for splice-modifying activity using a SMN2 minigene I believe, resulting in 52 compounds that they send Gabi and I for further experiments: We recieved the 52 compounds in a 96 well plate, and in subsequent plots we are naming them just by the well position they were shipped in (ie A01, A02, ..., E04). The molar concentration of each molecule was determined by Jingxin's lab and shipped at 500x concentration. I think molar concentration is based on EC90 of SMN2 splice modifying activity in their initial screen. We grew LCLs cells to a density of approximately 1M cells per mL, and split into seperate flasks each containing 7mL: 2 seperate replicate flasks for each molecule treatment, and 6 flasks of DMSO control (total of 110 samples), and applied 14uL of treatment. The following day, cells were collected by centrifugation, supernate was discarded and stored in 1mL of Trizol at -20C for futher processing. We extracted RNA by phase seperation (Trizol manufacturer's instructions). The top phase was mixed with an equal volume of ethanol, and applied to Zymo5 nucleic-acid purification columns. After two washes, we performed on-column DNAse I treatment, followed by an extra application of binding buffer and two more washes, and finally elution in water. Qubit quantification was used to quantify RNA before using [NEBNext ultra directional ii ](https://www.neb.com/products/e7760-nebnext-ultra-ii-directional-rna-library-prep-kit-for-illumina#Product%20Information) RNAseq kits with NEB's polyA capture kit. There are a couple optional steps in the kit preparation: We fragmented the RNA for only5 minutes, and performed size selection (step 6.3) to preferentially obtain longer insert sizes. The libraries were intially sequenced on a MiSeq (~20M reads total), to determine optimal re-pooling volumes (aiming for equal representation of each sample) and to confirm long insert sizes to justify the 150+150 paired end sequencing that we ended up doing on the NovaSeq S4 (~10B reads total). In practice, I noticed that one sample had very very low number of reads, and will likely have to be dropped from analysis (see below). Also, sample pooling volumes weren't as perfect as I would have hoped, but I think they are acceptable. see below.


In the remainder of this notebook, I will explore this data is ways similar to my previous notebooks regarding Jingxin's RNAseq - analyses that I can do quickly. The pre-processing that I have done leading up to this notebook includes read trimming ([fastp](https://academic.oup.com/bioinformatics/article/34/17/i884/5093234)), alignment with STAR (I used single pass mode, but perhaps before publication it might be worth aligning all samples and re-aligning samples for STAR's "two-pass" procedure for most sensitive identification of unannotated junctions), gene counting with featureCounts, differential expression analysis with edgeR (a separate contrast for each treatment, comparing to the same 6 DMSO replicates ), splicing quantification and differential testing with leafcutter (again, comparing each treatment to 6 DMSO replicates).

### Reads per sample

Let's just consider protein coding genes, as written out in my reformatted featureCounts output, which I will read in here. I will occasionally read in data from previous experiments too for comparison.

```{r}
# read in sample metadata


Metadata.ExpOf52 <- read_tsv("../code/config/samples.52MoleculeExperiment.tsv") %>%
  mutate(cell.type = "LCL", libType="polyA", rep=BioRep, old.sample.name=SampleID, dose.nM=NA) %>%
  mutate(treatment = if_else(Treatment=="DMSO", "DMSO", paste0("W", Treatment))) %>%
  dplyr::select(treatment, cell.type, dose.nM, libType, rep, old.sample.name) %>%
  mutate(SampleName = paste(treatment, dose.nM, cell.type, libType, rep, sep = "_"))

Metadata.PreviousExperiments <- read_tsv("../code/bigwigs/BigwigList.tsv",col_names = c("SampleName", "bigwig", "group", "strand")) %>%
  filter(strand==".") %>%
  dplyr::select(-strand) %>%
  mutate(old.sample.name = str_replace(bigwig, "/project2/yangili1/bjf79/20211209_JingxinRNAseq/code/bigwigs/unstranded/(.+?).bw", "\\1")) %>%
  separate(SampleName, into=c("treatment", "dose.nM", "cell.type", "libType", "rep"), convert=T, remove=F, sep="_") %>%
  left_join(
    read_tsv("../code/bigwigs/BigwigList.groups.tsv", col_names = c("group", "color", "bed", "supergroup")),
    by="group"
  )

FullMetadata <- bind_rows(Metadata.ExpOf52, Metadata.PreviousExperiments) %>%
  mutate(Experiment = case_when(
    cell.type == "Fibroblast" ~ "Single high dose fibroblast",
    startsWith(old.sample.name, "TitrationExp") ~ "Dose response titration",
    startsWith(old.sample.name, "chRNA") ~ "nascent RNA profiling",
    startsWith(old.sample.name, "NewMolecule") ~ "Single high dose LCL"
  )) %>%
  mutate(color = case_when(
    treatment == "DMSO" ~ "#969696",
    Experiment == "Single high dose LCL" ~ "#252525",
    TRUE ~ color
  )) %>%
  mutate(leafcutter.name = str_replace_all(old.sample.name, "-", "."))



gene.counts <- read_tsv("../code/DE_testing/ExpOf52_Counts.mat.txt.gz") %>%
  column_to_rownames("Geneid") %>%
  DGEList() %>%
  calcNormFactors()

counts.plot.dat <- gene.counts$samples %>%
  rownames_to_column("old.sample.name") %>%
  inner_join(FullMetadata, by="old.sample.name") %>%
  mutate(dose.nM = case_when(
    treatment == "DMSO" ~ "NA",
    cell.type == "Fibroblast" ~ "CC50 dose",
    Experiment == "Single high dose LCL" ~ "SMN_EC90 dose",
    TRUE ~ as.character(dose.nM)
  )) %>%
  mutate(label = dose.nM) %>%
  arrange(Experiment, treatment, dose.nM)

counts.plot.labels <- counts.plot.dat %>%
  dplyr::select(old.sample.name, label) %>% deframe()

counts.plot.dat %>%
  filter(Experiment=="Single high dose LCL")

ReadsPerDataset <- ggplot(counts.plot.dat, aes(x=old.sample.name, y=lib.size/2E6, fill=color)) +
  geom_col() +
  scale_fill_identity() +
  scale_x_discrete(name="dose (nanomolar)", label=counts.plot.labels) +
  scale_y_continuous(expand=c(0,0)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=3)) +
  theme(strip.text.x = element_text(size = 8)) +
  facet_grid(~Experiment, scales = "free_x", space = "free_x", labeller = label_wrap_gen(15)) +
  labs(title="RNA-seq datasets", y="Read count (M)")
ReadsPerDataset

```

So for the most part, these libraries were sequenced deeper than any of our previous experiments with Jingxin. Let's make note of that one outlier sample that was clearly not sequenced well. This sample will probably have to be excluded from further analysis...

```{r}
counts.plot.dat %>%
  arrange(lib.size) %>%
  head(1)

```

Ok so rep2 of the molecule in well C04 will probably have to be excluded. 4M reads isn't enough to get much out of, and I sort of worry whether those 4M reads are even from the corresponding library versus barcode contamination from other samples on the lane...

### Gene expression PCA

First let's perform PCA with all samples, including fibroblasts. I know this isn't necessarily the most interpretable, but I am just curious what the first few PCs will look like...


```{r}

SamplesToInclude <- FullMetadata %>%
  pull(old.sample.name)

CPM <- gene.counts %>%
  cpm(log=T, prior.count=T) %>%
  as.data.frame() %>%
  rownames_to_column("Geneid") %>%
  dplyr::select(Geneid, all_of(SamplesToInclude))

Top14K_ExpressedGenes <- (CPM %>%
  column_to_rownames("Geneid") %>%
  apply(1, mean) %>%
  sort(decreasing=T))[1:14000] %>%
  names()

pca.results <- CPM %>%
  filter(Geneid %in% Top14K_ExpressedGenes) %>%
  column_to_rownames("Geneid") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("old.sample.name") %>%
  dplyr::select(old.sample.name, PC1:PC6) %>%
  left_join(FullMetadata, by="old.sample.name")

ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nAll samples")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nAll samples")
ggplot(pca.results.to.plot, aes(x=PC5, y=PC6, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nAll samples")

```

Let's repeat but just include the dose titration experiment samples and this recent experiment of 52 molecules...


```{r}

SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL", "Dose response titration")) %>%
  pull(old.sample.name)

CPM <- gene.counts %>%
  cpm(log=T, prior.count=T) %>%
  as.data.frame() %>%
  rownames_to_column("Geneid") %>%
  dplyr::select(Geneid, all_of(SamplesToInclude))

Top14K_ExpressedGenes <- (CPM %>%
  column_to_rownames("Geneid") %>%
  apply(1, mean) %>%
  sort(decreasing=T))[1:14000] %>%
  names()

pca.results <- CPM %>%
  filter(Geneid %in% Top14K_ExpressedGenes) %>%
  column_to_rownames("Geneid") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("old.sample.name") %>%
  dplyr::select(old.sample.name, PC1:PC6) %>%
  left_join(FullMetadata, by="old.sample.name")

ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nTitration experiment and ExpOf52")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nTitration experiment and ExpOf52")
ggplot(pca.results.to.plot, aes(x=PC5, y=PC6, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nTitration experiment and ExpOf52")

pca.results.to.plot %>%
  filter(PC5>200)

```


Ok... some interesting patterns. That one outlier in PC5 is the sample with low read depth that I want to exclude.  It seems like the lab/batch effects are quite strong, the first two PCs have nothing to do with dose. (Lighter gray points are the DMSO controls in both experiments).. Maybe PC3 is related to this? But it's not going to be so straightforward to integrate this experiment of 52 data with the previous dose titration experiment also done in LCLs. Well at least previously when I perform PCA based on splicing junction excision ratios these batch and tissue effects seem to dissappear in the PCA, so maybe those splicing comparisons might be more fair without more careful integration. Now let's try PCA again but now just looking at the samples in this recent experiment of 52 molecules, to check at least that the DMSO samples cluster together...

```{r}
SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(old.sample.name)

CPM <- gene.counts %>%
  cpm(log=T, prior.count=T) %>%
  as.data.frame() %>%
  rownames_to_column("Geneid") %>%
  dplyr::select(Geneid, all_of(SamplesToInclude))

Top14K_ExpressedGenes <- (CPM %>%
  column_to_rownames("Geneid") %>%
  apply(1, mean) %>%
  sort(decreasing=T))[1:14000] %>%
  names()

pca.results <- CPM %>%
  filter(Geneid %in% Top14K_ExpressedGenes) %>%
  column_to_rownames("Geneid") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("old.sample.name") %>%
  dplyr::select(old.sample.name, PC1:PC6) %>%
  left_join(FullMetadata, by="old.sample.name")

ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nExpOf52")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nExpOf52")
ggplot(pca.results.to.plot, aes(x=PC5, y=PC6, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nExpOf52")

```

Ok, even though I am looking at only the first 6 PCs, it seems pretty clear that the DMSO samples don't cleanly cluster together. Perhaps some of these molecules weren't very active.

Now just for curiosity, let's just include the fibroblast and the dose titration experiment.

```{r}
SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Dose response titration", "Single high dose fibroblast")) %>%
  pull(old.sample.name)

CPM <- gene.counts %>%
  cpm(log=T, prior.count=T) %>%
  as.data.frame() %>%
  rownames_to_column("Geneid") %>%
  dplyr::select(Geneid, all_of(SamplesToInclude))

Top14K_ExpressedGenes <- (CPM %>%
  column_to_rownames("Geneid") %>%
  apply(1, mean) %>%
  sort(decreasing=T))[1:14000] %>%
  names()

pca.results <- CPM %>%
  filter(Geneid %in% Top14K_ExpressedGenes) %>%
  column_to_rownames("Geneid") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("old.sample.name") %>%
  dplyr::select(old.sample.name, PC1:PC6) %>%
  left_join(FullMetadata, by="old.sample.name")

ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nTitrationExp and FibroblastExp")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nTitrationExp and FibroblastExp")
ggplot(pca.results.to.plot, aes(x=PC5, y=PC6, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using top 14K genes\nTitrationExp and FibroblastExp")
```

Wow, so in that, the second PC seems to pretty obviously reflect dose. I am quite surprised we don't see something so obvious with the new samples... Somewhat concerning. Of course, perhaps it's not too surprising considering this experiment just has so many more samples, only a few of which are DMSO, that maybe I shouldn't expect some dose effect in the first few PCs... One other thing I have done that was helpful last time was to perform PCA on the titration series then project new samples in the same PC space. I'll try that later...


Now Let's check for some splicing effects. Maybe that will be more illuminating...

### Verify genome-wide GA-GT splice site induction

Let's start by checking for global enrichment of GA|GT juncs... If this doesn't show an effect I'll be very skeptical about the usefulness of this data.

```{r}
leafcutter.sig <- Sys.glob("../code/SplicingAnalysis/leafcutter/differential_splicing/ExpOf52_*_cluster_significance.txt") %>%
  setNames(str_replace(., "../code/SplicingAnalysis/leafcutter/differential_splicing/ExpOf52_(.+?)_cluster_significance.txt", "W\\1")) %>%
  lapply(fread) %>%
  bind_rows(.id="treatment")

leafcutter.effects <- Sys.glob("../code/SplicingAnalysis/leafcutter/differential_splicing/ExpOf52_*_effect_sizes.txt") %>%
  setNames(str_replace(., "../code/SplicingAnalysis/leafcutter/differential_splicing/ExpOf52_(.+?)_effect_sizes.txt", "W\\1")) %>%
  lapply(fread, col.names=c("intron", "logef", "treatment_PSI", "DMSO_PSI", "deltapsi")) %>%
  bind_rows(.id="treatment") %>%
  mutate(name = str_replace(intron, "(.+?):(.+?):(.+?):clu.+([+-])", "\\1_\\2_\\3_\\4")) %>%
  mutate(deltapsi = deltapsi * -1, logef=logef*-1) %>%
  mutate(cluster = str_replace(intron, "(.+?):.+?:.+?:(clu.+[+-])", "\\1:\\2"))


Introns <- read_tsv("../code/SplicingAnalysis/FullSpliceSiteAnnotations/JuncfilesMerged.annotated.basic.bed.5ss.tab.gz", col_names = c("name", "seq", "score")) %>%
  separate(name, into=c("name", "pos"), sep = "::")

Introns.annotations <- read_tsv("../code/SplicingAnalysis/FullSpliceSiteAnnotations/JuncfilesMerged.annotated.basic.bed.gz") %>%
  mutate(name = paste(chrom, start, end, strand, sep = "_")) %>%
  left_join(Introns, by="name")


leafcutter.effects %>%
  inner_join(
    Introns.annotations %>%
    mutate(IsGAGT = if_else(str_detect(seq, "^\\w\\wGAGT"), "GA-GT", "not GA-GT")),
    by="name") %>%
  # count(IsGAGT, treatment)
  ggplot(aes(x=logef, group=treatment)) +
  stat_ecdf() +
  coord_cartesian(xlim=c(-2,2)) +
  facet_wrap(~IsGAGT) +
  theme_bw() +
  labs(title="Distribution of log(FC) for all tested introns\nExpOf52", y="ecdf", x="Splicing log(FC)")

```


Ok so I see some subtle general up-regulation of GA-GT introns... That's reassuring. But the degree to which all samples have about the same general effect size is surprising to me... I expected some of the compounds to be less active... Even though we normalized concentration based on SMN2 minigene EC90, I suspected a bit of noise in that estimate anyway. Let's compare this to the analogous plot from the previous fibroblast dataset where I also did differential splicing with leafcutter to estimate splicing log(FC)...

```{r}
leafcutter.effects.fibroblasts <- list.files(path="../code/SplicingAnalysis/leafcutter/differential_splicing", pattern="^[A-Z]+[0-9]*_effect_sizes.txt", full.names=T) %>%
  setNames(str_replace(., "../code/SplicingAnalysis/leafcutter/differential_splicing/(.+?)_effect_sizes.txt", "\\1")) %>%
  lapply(fread, col.names=c("intron", "logef", "treatment_PSI", "DMSO_PSI", "deltapsi")) %>%
  bind_rows(.id="treatment") %>%
  mutate(name = str_replace(intron, "(.+?):(.+?):(.+?):clu.+([+-])", "\\1_\\2_\\3_\\4")) %>%
  mutate(deltapsi = deltapsi * -1, logef=logef*-1) %>%
  mutate(cluster = str_replace(intron, "(.+?):.+?:.+?:(clu.+[+-])", "\\1:\\2"))

leafcutter.sig.fibroblasts <-list.files(path="../code/SplicingAnalysis/leafcutter/differential_splicing", pattern="^[A-Z]+[0-9]*_cluster_significance.txt", full.names=T) %>%
  setNames(str_replace(., "../code/SplicingAnalysis/leafcutter/differential_splicing/(.+?)_cluster_significance.txt", "\\1")) %>%
  lapply(fread) %>%
  bind_rows(.id="treatment")


Fibroblast.colors <- FullMetadata %>%
  filter(cell.type=="Fibroblast") %>%
  distinct(treatment, .keep_all=T) %>%
  dplyr::select(treatment, color) %>%
  deframe()

leafcutter.effects.fibroblasts %>%
  inner_join(
    Introns.annotations %>%
    mutate(IsGAGT = if_else(str_detect(seq, "^\\w\\wGAGT"), "GA-GT", "not GA-GT")),
    by="name") %>%
  # count(IsGAGT, treatment)
  ggplot(aes(x=logef, color=treatment)) +
  stat_ecdf() +
  scale_color_manual(values=Fibroblast.colors) +
  coord_cartesian(xlim=c(-2,2)) +
  facet_wrap(~IsGAGT) +
  theme_bw() +
  labs(title="Distribution of log(FC) for all tested introns\nFibroblast Exp", y="ecdf", x="Splicing log(FC)")


```
Ok, so maybe the degree of these 52 treatments are roughly on the order of the CUOME treatment in fibroblast data. Not too strong. I'm still not totally sure this is even real... 

Ok, so now let's doing PCA just using GA|GT juncs... I think this should really work. Let's start by including all samples. If I don't really see DMSO control pop out from the rest, I will be a bit worried.


### Splicing correlation matrix and PCA


Actually before doing PCA, to get a broad overview of results, let's calculate the correlation coefficient across samples based on splicing. For sake of quickly getting an overview, and picking reasonable GA-GT introns (that aren't too noisy to measure) let's just consider the 500ish GA-GT introns I previously modelled. These are a pretty 'good' set of differentially spliced GA-GT introns based on risdiplam,branaplam,C2C5...



```{r}
#Read in splice junction count table

PSI <- fread("../code/SplicingAnalysis/leafcutter_all_samples/leafcutter_perind_numers.bed.gz", sep = '\t' ) %>%
  dplyr::select(-"NewMolecule.C04.2")

Modelled.introns <- read_tsv("../output/EC50Estimtes.FromPSI.txt.gz") %>%
  dplyr::select(`#Chrom`, start, end, strand=strand.y)


colnames(PSI)

PSI.GAGT.Only <- PSI %>%
  inner_join(Modelled.introns) %>%
  dplyr::select(junc, A10.1:NewMolecule.E07.6) %>%
  drop_na()


PSI.GAGT.Only.cormat <- PSI.GAGT.Only %>%
  column_to_rownames("junc") %>%
  cor(method='s')

heatmap.2(PSI.GAGT.Only.cormat, trace='none')
```

And now let's do PCA based on these introns...

```{r}
pca.results <- PSI.GAGT.Only %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nAll samples")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nAll samples")
ggplot(pca.results.to.plot, aes(x=PC5, y=PC6, color=color, shape=Experiment)) +
  geom_point(size=3) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nAll samples")


```

That is actually kind of reassuring. The first PC is definitely related to dose. And all DMSO samples cluster together. However, again, the effective dosage seems a bit weak compared to fibroblast experiments... Looks like most of the samples are roughly equivalent to ~100nM risdiplam, so still perhaps on the high end of what would be realistic clinical concentration but I think a slightly higher dose would have been more informative. In this particular PCA, it looks like PC3 sepearates branaplam vs risdiplam, and it's not totally clear how these molecules are different along that axis... I could try, as mentioned before just doing PCA using the dose titration experiments then projecting the new samples in that space... But first let's do a PCA just using the samples from the new experiment..

```{r}
SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)

pca.results <- PSI.GAGT.Only %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nExpOf52 samples")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nExpOf52 samples")
ggplot(pca.results.to.plot, aes(x=PC5, y=PC6, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nExpOf52 samples")
```

Ok, reassuringly, again DMSO samples stick out in the first PC, and I see a few biological replicates are close together.

Ok, now I want to try projecting the old titration series data into this same PCA spaace, which might help interpretability.

```{r}
SamplesToProject <- FullMetadata %>%
  filter(!Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)

pca.results.to.plot_AdditionProjections <- PSI.GAGT.Only %>%
  dplyr::select(junc, all_of(SamplesToProject)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% predict(pca.results, .) %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")


TreatmentColorsForLabels.FibroblastColorsAdded <-
FullMetadata %>%
  group_by(treatment) %>%
  filter(dose.nM == max(dose.nM) | treatment == "DMSO" | cell.type=="Fibroblast") %>%
  ungroup() %>%
  distinct(treatment, .keep_all=T) %>%
  arrange(dose.nM) %>%
  mutate(vjust=row_number()*1.2)

TreatmentColorsLabels.Layer.FibroblastColorsAdded <- geom_text(
    data = TreatmentColorsForLabels.FibroblastColorsAdded, 
    aes(label=treatment, color=color, vjust=vjust),
    y=Inf, x=Inf, hjust=1.05
  )

  
ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nExpOf52 samples", caption="Additional samples projected")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 1090 GAGT introns\nAll samples")

```

I think these PCA are interseting... it's still hard to make sense of it all... And this isn't really the way to identify new axes of specifity since the introns were selected precisely because of their effect in risdiplam/branaplam/C2C5 in previous experiments... Better would be to identify all differentially spliced introns and perform a PCA on that, also projecting the old risdiplam/branaplam/C2C5 data in the PCA for help with interpretation... Let's work towards that....


### Leafcutter differential splicing

First, let's explore the leafcutter differential splicing results... From previous plots, I suspect the effective dosage of these new data to be much less than the fibroblast data, and perhaps we will get much fewer significant differential splicing events... First Let's check it out, the number of differentially spliced clusters, some P value histograms, etc.

```{r}
leafcutter.sig %>%
  sample_n_of(12, treatment) %>%
  filter(!is.na(p)) %>%
  ggplot(aes(x=p)) +
    geom_histogram() +
    facet_wrap(~treatment, scales="free") +
  labs(caption="P value histograms for some random splicing contrasts", title="Exp of 52")
  
```

...and compare that to the fibroblast tests...

```{r}
leafcutter.sig.fibroblasts %>%
  filter(!is.na(p)) %>%
  ggplot(aes(x=p)) +
    geom_histogram() +
    facet_wrap(~treatment, scales="free") +
  labs(caption="P value histograms for some random splicing contrasts", title="fibroblast experiment")
```

ok, so a lot these will probably have even less significant hits than the fibroblast CUOME contrast.

Let's see how many hits at an FDR of 10%...


```{r}
bind_rows(
bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig) %>%
  filter(p.adjust < 0.01) %>%
  count(treatment) %>%
  mutate(FDR = 0.01),
bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig) %>%
  filter(p.adjust < 0.05) %>%
  count(treatment) %>%
  mutate(FDR = 0.05),
bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig) %>%
  filter(p.adjust < 0.1) %>%
  count(treatment) %>%
  mutate(FDR = 0.1)) %>%
  left_join(FullMetadata, by='treatment') %>%
  ggplot(aes(x=treatment, y=n)) +
  geom_bar(stat="identity",position = "identity", alpha=.15) +
  facet_grid(~Experiment, scales = "free", space = "free_x", labeller = label_wrap_gen(10)) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=6)) +
  theme(strip.text.x = element_text(size = 5)) +
  labs(title="Number of leafcutter signigicant clusters", x="treatment", caption="FDR 1%, 5%, 10%")

bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig) %>%
  filter(p.adjust < 2) %>%
  count(treatment) %>%
  left_join(FullMetadata, by='treatment') %>%
  ggplot(aes(x=treatment, y=n)) +
  geom_bar(stat="identity",position = "identity", alpha=1) +
  facet_grid(~Experiment, scales = "free", space = "free_x", labeller = label_wrap_gen(10)) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=6)) +
  theme(strip.text.x = element_text(size = 5)) +
  labs(title="Number of leafcutter tests", x="treatment")

```

Ok, most treatments have even less hits than the CUOME contrast. In principle, sharing information across samples could help with this. I think this might be important towards finding difference amongst samples and identification of new specificities. Like mash could be useful. But that takes a bit of time to implement... leafcutter doesn't just output standard errors, which I think are required. I suppose one hackish solution would be to perform a contrast with all treatments combined, vs DMSO... then identify significant clusters. To find more molecule-specific effects, I could then look at the nominal P value from the non aggregated treatment contrast.

Following this idea, I did a contrast with all molecules vs DMSO... Here are the leafcutter results...


```{r}
leafcutter.effects.MergedContrast <- read_tsv("../code/SplicingAnalysis/MergedExp52_Contrast/_effect_sizes.txt")
leafcutter.sig.MergedContrast <- read_tsv("../code/SplicingAnalysis/MergedExp52_Contrast/_cluster_significance.txt")

bind_rows(
bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig, leafcutter.sig.MergedContrast) %>%
  filter(p.adjust < 0.01) %>%
  count(treatment) %>%
  mutate(FDR = 0.01),
bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig, leafcutter.sig.MergedContrast) %>%
  filter(p.adjust < 0.05) %>%
  count(treatment) %>%
  mutate(FDR = 0.05),
bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig, leafcutter.sig.MergedContrast) %>%
  filter(p.adjust < 0.1) %>%
  count(treatment) %>%
  mutate(FDR = 0.1)) %>%
  left_join(FullMetadata, by='treatment') %>%
  replace_na(list(treatment="Merged", Experiment="Single high dose LCL")) %>%
  ggplot(aes(x=treatment, y=n)) +
  geom_bar(stat="identity",position = "identity", alpha=.15) +
  facet_grid(~Experiment, scales = "free", space = "free_x", labeller = label_wrap_gen(10)) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=6)) +
  theme(strip.text.x = element_text(size = 5)) +
  labs(title="Number of leafcutter signigicant clusters", x="treatment", caption="FDR 1%, 5%, 10%")

bind_rows(leafcutter.sig.fibroblasts, leafcutter.sig, leafcutter.sig.MergedContrast) %>%
  filter(p.adjust < 2) %>%
  count(treatment) %>%
  left_join(FullMetadata, by='treatment') %>%
  replace_na(list(treatment="Merged", Experiment="Single high dose LCL")) %>%
  ggplot(aes(x=treatment, y=n)) +
  geom_bar(stat="identity",position = "identity", alpha=1) +
  facet_grid(~Experiment, scales = "free", space = "free_x", labeller = label_wrap_gen(10)) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=6)) +
  theme(strip.text.x = element_text(size = 5)) +
  labs(title="Number of leafcutter tests", x="treatment")


```

Wow. That did surprisingly little. I don't have the time to troubleshoot if there is a bug or if this is a legitamite result of doing this merged contrast. 

So now let's try counting how many unique clusters there are amongst all the attempted contrasts. Let's be permissive and consider FDR 10%

```{r}
#Number of total hits (not necessarily unique)
leafcutter.sig %>%
  filter(p.adjust < 0.1) %>% nrow()

#Number of unique cluster hits amongst all hits
leafcutter.sig %>%
  filter(p.adjust < 0.1) %>%
  distinct(cluster) %>% nrow()
```

Ok, so a lot of the hits are unique to some contrasts, in the sense that I get 10803/32162 *unique* clusters with FDR<10%. I think what I will do next is do the PCA, but just include introns that are FDR<10% in at least one contrast, and also consider some minimal effect size, and also consider just GAGT introns within those clusters.

```{r}
IntronsToInclude <- leafcutter.sig %>%
  filter(p.adjust < 0.1) %>%
  inner_join(leafcutter.effects, by=c("treatment", "cluster")) %>%
  filter(deltapsi > 0.01) %>%
  inner_join(Introns.annotations, by='name') %>%
  filter(str_detect(seq, "^\\w\\wGAGT")) %>%
  pull(intron)



length(IntronsToInclude)

PSI.IntronsToInclude <- PSI %>%
  filter(junc %in% IntronsToInclude) %>%
  dplyr::select(junc, A10.1:NewMolecule.E07.6) %>%
  drop_na()

dim(PSI.IntronsToInclude)

```
Ok, so only 475 introns left... This probably won't change anything from my previous PCA with ~1000 GAGT introns... Well let's follow through anyway...


```{r}

SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)


pca.results <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)

summary(pca.results)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

SamplesToProject <- FullMetadata %>%
  filter(!Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)

pca.results.to.plot_AdditionProjections <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToProject)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% predict(pca.results, .) %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")


TreatmentColorsForLabels.FibroblastColorsAdded <-
FullMetadata %>%
  group_by(treatment) %>%
  filter(dose.nM == max(dose.nM) | treatment == "DMSO" | cell.type=="Fibroblast") %>%
  ungroup() %>%
  distinct(treatment, .keep_all=T) %>%
  arrange(dose.nM) %>%
  mutate(vjust=row_number()*1.2)

TreatmentColorsLabels.Layer.FibroblastColorsAdded <- geom_text(
    data = TreatmentColorsForLabels.FibroblastColorsAdded, 
    aes(label=treatment, color=color, vjust=vjust),
    y=Inf, x=Inf, hjust=1.05
  )

  
ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 475 GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")
ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 475 GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")

```

Ok this is actually somewhat interpretable. Again the main axis of specificity seems to be along branaplam vs risdiplam effects, and there is plenty of stuff in between. WD07 is actually 500nM risdiplam, which looks fitting in relation to previous experiments. I think perhaps WC05 (SMSM32) and WD08 (4F-Py) the most interesting point I see so far... as both replicates doesn't neatly lie on either the risdiplam axis or the branaplam axis. Just browsing these molecules structure that ZT shared with me, these seem by my eye pretty structurally distinct. I would plot the structures of these in the PCA but `ggimage` is kind of tricky to install in my current R environment, so that may have to wait for later

- update: I turned out making that plot with ggimage in a seperate notebook that I run in a different R environment... I wrote out the data for that plot in the block below and wrote a seperate notebook to just read in that data and plot...


```{r, eval=F}
FullMetadata %>%
  write_tsv("../output/20230223_FullMetadata.tsv")


bind_rows(pca.results.to.plot,
          pca.results.to.plot_AdditionProjections) %>%
  write_tsv("../output/20230223_PCA_Projections.tsv.gz")
 

PSI.IntronsToInclude %>%
  write_tsv("../output/20230223_IntronsToInclude.tsv.gz")

```


Let's make sure these three molecules don't have very different read depth or something...

```{r}
pca.results.to.plot %>%
  inner_join(
    counts.plot.dat %>%
      dplyr::select(SampleName, lib.size)
    ) %>%
ggplot( aes(x=PC1, y=PC2, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment, color=lib.size)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  scale_color_viridis_c(trans='log10') +
  # scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 475 GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")
```
Ok, that makes me feel better these three interesting candidate molecules aren't just outliers in read depth.

Next I want to plot a heatmap of the correlation matrix based on these introns, I wonder if any clear clustering patterns will come out of this... Part of what I want to get out of this is to answer if these three outliers (WD08, WC05, WB12) are similar to each other, or if they are as distant from eachother as they are from risdiplam/branaplam.

```{r}

SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL", "Dose response titration")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)

PSI.IntronsToInclude.cormat <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  cor(method='s')

GetColorPallete <- function(MyVector, PalletteName = "Dark2"){
  UniqueFactors <- MyVector %>% unique()
  return (setNames(brewer.pal(n=length(UniqueFactors), name=PalletteName), UniqueFactors))
}



MetadataForCormat <- data.frame(leafcutter.name = PSI.IntronsToInclude.cormat %>%
  rownames()) %>%
  left_join(FullMetadata) %>%
  mutate(color = if_else(
    treatment %in% c("WD08", "WC05", "WB12"),
    "#e34a33",
    color
  )) %>%
  mutate(ExperimentColor = recode(Experiment, !!!GetColorPallete(FullMetadata$Experiment)
))


  
heatmap.2(PSI.IntronsToInclude.cormat, trace='none', RowSideColors = MetadataForCormat$color, ColSideColors = MetadataForCormat$color, labRow = MetadataForCormat$treatment, cexRow=0.2, labCol = MetadataForCormat$treatment, cexCol=0.2)

```

```{r, eval=F}
pdf("../code/scratch/Heatmap.pdf")
heatmap.2(PSI.IntronsToInclude.cormat, trace='none', RowSideColors = MetadataForCormat$color, ColSideColors = MetadataForCormat$color, labRow = MetadataForCormat$treatment, cexRow=0.2, labCol = MetadataForCormat$treatment, cexCol=0.2)
dev.off()
```


Ok samples mostly just are clustering by effective dose, so you can't really see the specificity effect here like you can in the second PC above... Well, you can see in the hierarchal clustering that both reps of WC05, a rep of WD08, and a rep of WB12 cluster together... And far away on another branch are the other replicates of WD08 and WB12...

so maybe their effects are similar? It's just really hard to tell.

As a slightly less biased way of making this PCA and possibly reproducing the potential uniqueness of these 3 molecules, let's try changing the criteria for including introns to be more inclusive.

```{r}
IntronsToInclude <- leafcutter.sig %>%
  filter(p.adjust < 0.1) %>%
  inner_join(leafcutter.effects, by=c("treatment", "cluster")) %>%
  filter(abs(deltapsi) > 0.01) %>%
  distinct(name, .keep_all=T) %>%
  inner_join(Introns.annotations, by='name') %>%
  # filter(str_detect(seq, "^\\w\\wGAGT")) %>%
  pull(intron)

length(IntronsToInclude)

PSI.IntronsToInclude <- PSI %>%
  filter(junc %in% IntronsToInclude) %>%
  dplyr::select(junc, A10.1:NewMolecule.E07.6) %>%
  drop_na()

dim(PSI.IntronsToInclude)

SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)


pca.results <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name") %>%
  mutate(color = if_else(
    treatment %in% c("WD08", "WC05", "WB12"),
    "#e34a33",
    color
  )) 

SamplesToProject <- FullMetadata %>%
  filter(!Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)

pca.results.to.plot_AdditionProjections <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToProject)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% predict(pca.results, .) %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

  
ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 23416 introns\nAll samples", caption="Introns based on being significant in any contrast")

pca.results.to.plot %>%
  inner_join(
    counts.plot.dat %>%
      dplyr::select(SampleName, lib.size)
    ) %>%
ggplot(aes(x=PC1, y=PC2, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment, color=lib.size)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  scale_color_viridis_c(trans='log10') +
  # scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 23416 introns\nAll samples", caption="Introns based on being significant in any contrast. Colored by depth")

ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 23416 introns\nAll samples", caption="Introns based on being significant in any contrast")

ggplot(pca.results.to.plot, aes(x=PC4, y=PC5, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 23416 introns\nAll samples", caption="Introns based on being significant in any contrast")
```
Ok interesting... when you look at all significant introns with abs(deltapsi) > 0.01, the first few PCs measure hidden technical variables that don't seem to cluster biological replicates very well. deeper PCs seem to work better to find the dosage axis (PC4) branaplam vs risdiplam axis (PC5) and biological replicates still are close to one another... But in this space, a different set of molecules starts to show up, and perhaps on a different 'arm' of the data. I still see D08 sticking out.. WE02 also sticks out here along that axis, and WE02 was previously along the branaplam axis and is structurally most like branaplam.

Now I'm a bit concerned about the robustness of the PCA to interpret it as being a distinct thing.. Let's try changing up the intron critera a tad a again and see how robust the PCA is...

```{r}
IntronsToInclude <- leafcutter.sig %>%
  filter(p.adjust < 0.1) %>%
  inner_join(leafcutter.effects, by=c("treatment", "cluster")) %>%
  filter(logef > 0) %>%
  distinct(name, .keep_all=T) %>%
  inner_join(Introns.annotations, by='name') %>%
  filter(str_detect(seq, "^\\w\\wGAGT")) %>%
  pull(intron)

length(IntronsToInclude)

PSI.IntronsToInclude <- PSI %>%
  filter(junc %in% IntronsToInclude) %>%
  dplyr::select(junc, A10.1:NewMolecule.E07.6) %>%
  drop_na()

dim(PSI.IntronsToInclude)

SamplesToInclude <- FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)


pca.results <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name") %>%
  mutate(color = if_else(
    treatment %in% c("WD08", "WC05", "WB12"),
    "#e34a33",
    color
  )) 

SamplesToProject <- FullMetadata %>%
  filter(!Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  pull(leafcutter.name)

pca.results.to.plot_AdditionProjections <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToProject)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% predict(pca.results, .) %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

  
ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")

pca.results.to.plot %>%
  inner_join(
    counts.plot.dat %>%
      dplyr::select(SampleName, lib.size)
    ) %>%
ggplot(aes(x=PC1, y=PC2, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment, color=lib.size)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  scale_color_viridis_c(trans='log10') +
  # scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Introns based on being significant in any contrast. Colored by depth")

ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")

ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  # geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")

ggplot(pca.results.to.plot, aes(x=PC4, y=PC5, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")

ggplot(pca.results.to.plot, aes(x=PC4, y=PC5, color=color, shape=Experiment)) +
  geom_text(size=3, aes(label=treatment)) +
  # geom_point(data = pca.results.to.plot_AdditionProjections) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")
```



Hmm, ok much like the first PCA. Of course these are surely mostly overlapping introns. But at least is sort of comforting. I thought looking at deeper PCs might reveal something... Mostly reveals that interpreting PCs is not always straightforward.

Also, if C05, D08, and B12 have unique properties, it is not so clear if  are similar or distinct from each other.

First let's write out these PCA, for later use...

```{r, eval=F}
pca.results.to.plot %>%
  write_tsv("../output/20230227_PCA_Projections.tsv.gz")
  
```


## Investigation of the C05, D08, and B12

Let's check the behavior of the hits (and all samples) on some particular introns or genes... The HTT, MYB, SMN2, STAT1

First let's glance at the differential expression test results (which also include effect size estimates for these genes). Since gene expression changes are generally less noisy to estimate than splicing, we can also use these as proxies for risdiplam vs branaplam-specific effects for a set of known genes.

```{r}
diff.expression <- fread("../code/DE_testing/ExpOf52_Results.txt.gz") %>%
  mutate(treatment = str_replace(treatment, "group(.+?)", "W\\1")) %>%
  separate(Geneid, into=c("EnsemblID", "symbol"), sep='_')

diff.expression %>%
  filter(symbol %in% c("STAT1", "HTT", "MYB", "SMN2", "SMN1", "GALC", "FOXM1")) %>%
  left_join(FullMetadata) %>%
  mutate(color = if_else(
    treatment %in% c("WD08", "WC05", "WB12"),
    "#e34a33",
    color
  )) %>%
  mutate(Sig = case_when(
    PValue < 0.001 ~ "!",
    PValue < 0.05 ~ "*",
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x=treatment, y=logFC, fill=color)) +
  geom_col() +
  geom_text(aes(label=Sig), y=Inf, vjust=1) +
  scale_fill_identity() +
  facet_wrap(~symbol) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=3))
```


Hmm.. Offhand I can't really remember which if any of these genes (besides HTT) should be branaplam or risdiplam-specific. I think one helpful way to plot this will be to incorporate titration series data somehow, but that is a bit of work. Another way will be to color all samples (some of which are clearly on the risdiplam or branaplam axis) by the degree to which they were on those axes in PC space. Let's come up with a color scheme based on that idea...

For example, looking at the previous PCA plot...

If PC2 > 0 then blue,
ifelse PC2 < 0, then green
And PC2 is an alpha fill scale.

```{r, fig.width=8, fig.height=6}

NewColors <- pca.results.to.plot %>%
  filter(Experiment == "Single high dose LCL") %>%
  group_by(treatment) %>%
  summarise(PC1 = mean(PC1), PC2=mean(PC2)) %>%
  mutate(NewColor = case_when(
    PC2 > 0 ~ "#08519c",
    PC2 < 0 ~ "#006d2c",
  ))

MinPC1 <- min(NewColors$PC1)
MaxPC1 <- max(NewColors$PC1)

NewColors <- NewColors %>%
  mutate(alpha = (PC1 - MinPC1)/(MaxPC1 - MinPC1)) %>%
  mutate(NewColor = if_else(
    treatment %in% c("WD08", "WC05", "WB12"),
    "#e34a33",
    NewColor
  ))

diff.expression %>%
  filter(symbol %in% c("STAT1", "HTT", "MYB", "SMN2", "SMN1", "GALC", "FOXM1")) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  mutate(Sig = case_when(
    PValue < 0.001 ~ "!",
    PValue < 0.05 ~ "*",
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x=treatment, y=logFC, fill=NewColor)) +
  geom_col(aes(alpha=alpha)) +
  geom_text(aes(label=Sig), y=Inf, vjust=1) +
  scale_fill_identity() +
  facet_wrap(~symbol) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=6), legend.position='none') +
  labs(title="Expression changes in genes of interest", y="log2FC", caption=str_wrap("P<0.05 = *\nP<0.001 = !\nColor and  fill transparency based on PC1 & PC2 to distinguish risdiplam/branaplam axes", 30))

```

```{r, eval=F}
ggsave("../code/scratch/scratch.pdf", height=6, width=10)
```


Ok, so based on these subset of genes, the red molecules have some risdiplam-like effects and some branaplam-like effects. Let's look more globally at the DE results..

```{r}

XaxisColors <- NewColors %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  arrange(treatment) %>% pull(NewColor)


NumberDE.genes <- bind_rows(
diff.expression %>%
  filter(FDR < 0.01) %>%
  count(treatment) %>%
  mutate(FDR = 0.01),
diff.expression %>%
  filter(FDR < 0.05) %>%
  count(treatment) %>%
  mutate(FDR = 0.05),
diff.expression %>%
  filter(FDR < 0.1) %>%
  count(treatment) %>%
  mutate(FDR = 0.1)) %>%
  left_join(NewColors, by='treatment') %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(x=treatment, y=n)) +
  geom_bar(stat="identity",position = "identity", alpha=.15) +
  geom_text(y=Inf,  vjust=1, label="*", aes(color=NewColor, alpha=alpha)) +
  Rotate_x_labels +
  scale_color_identity() +
  theme(axis.text.x = element_text(size=6)) +
  theme(legend.position='none') +
  labs(title="Number of DE genes", x="treatment", caption=str_wrap("FDR 1%, 5%, 10%\nAsterisk Color for PC1/PC2 to approximate branaplam/risdiplam axes", 30))
NumberDE.genes
```

Ok wow WD08 is a real outlier there. Tons of DE genes... It also had the most leafcutter hits from this experiment. I'll have to look more into that later. First let's zoom in on the y=axis to better look at the other contrasts.

```{r}
NumberDE.genes +
  coord_cartesian(ylim=c(0,20))
```

Ok so a typical contrast has more like 5-10 DE genes at a reasonable FDR threshold. It's also worth noting that the branaplam like molecules, when roughly have similar amounts of differential splicing, have much less differential expression.

Let me go back to WD08... I'm wondering if this sample just had a crazy amount of read depth or something..

```{r}

counts.plot.dat %>%
  filter(Experiment == "Single high dose LCL") %>%
  dplyr::select(SampleName, lib.size, treatment) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(SampleName = fct_reorder(SampleName, PC2 * alpha)) %>%
  ggplot(aes(x=SampleName, y=lib.size)) +
  geom_col(aes(fill=NewColor, alpha=alpha)) +
  Rotate_x_labels +
  scale_fill_identity() +
  theme(axis.text.x = element_text(size=6)) +
  theme(legend.position='none')


```
Ok, so those three treatments, included WD08 are not huge outliers in terms of sequencing depth. Maybe there is something genuinely different about it that makes it have many DE genes and splicing changes.

Let's go back to looking at genomewide GA-GT activation levels...

```{r}

leafcutter.effects %>%
  inner_join(
    Introns.annotations %>%
    mutate(IsGAGT = if_else(str_detect(seq, "^\\w\\wGAGT"), "GA-GT", "not GA-GT")),
    by="name") %>%
  group_by(IsGAGT, treatment) %>%
  summarise(medLogef = median(logef)) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(x=treatment, y=medLogef, fill=NewColor, alpha=alpha)) +
  geom_col() +
  scale_fill_identity() +
  coord_flip() +
  facet_wrap(~IsGAGT) +
  theme(axis.text.y = element_text(size=6)) +
  theme(legend.position='none') +
  labs(title="3 treatments in question still activate GA-GT\nto similar levels as branaplam/risdiplam -like treatments", y="median splicing logFC")

```

Ok I have to do this for all dinucleotides... 

```{r, fig.width=7, fig.height=10}

leafcutter.effects %>%
  inner_join(
    Introns.annotations %>%
    mutate(UpstreamOfDonor2BaseSeq = substr(seq, 3, 4)),
    by="name") %>%
  group_by(UpstreamOfDonor2BaseSeq, treatment) %>%
  summarise(medLogef = median(logef)) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(y=treatment, x=UpstreamOfDonor2BaseSeq, fill=medLogef)) +
  geom_raster() +
  scale_fill_viridis_c() +
  theme(axis.text.y = element_text(size=6, colour = XaxisColors))
```

Ok, well firstly it is nice to see preferential activation of GA across the board. And there isn't any other dinucleotide that pops out in the three red samples in question. And the degree of GA-GT activation doesn't seem all that different from other treatments similar in PC1. I think this is particularly interesting for WD08 which at 3-4x as many differentially spliced clusters and ~100x as many DE genes as a typical treatment. I'm leaning towards thinking this molecule just might elicit non-splicing related toxic effects. Let's count the number of differentially spliced NNGT introns for each effect to better understand this... I'm wondering whether those 3-4x as many differentially spliced introns are from non GAGT introns (eg indirect effects).



```{r}

# leafcutter.effects %>%
#   inner_join(
#     Introns.annotations %>%
#     mutate(UpstreamOfDonor2BaseSeq = substr(seq, 3, 4)),
#     by="name") %>%
#   group_by(UpstreamOfDonor2BaseSeq, treatment) %>%
#   summarise(medLogef = median(logef)) %>%
#   left_join(NewColors) %>%
#   mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
#   mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
#   ggplot(aes(y=treatment, x=UpstreamOfDonor2BaseSeq, fill=medLogef)) +
#   geom_raster() +
#   scale_fill_viridis_c() +
#   theme(axis.text.y = element_text(size=6, colour = XaxisColors))

leafcutter.effects_and_sig <- leafcutter.effects %>%
  inner_join(leafcutter.sig) %>%
  inner_join(
    Introns.annotations %>%
    mutate(UpstreamOfDonor2BaseSeq = substr(seq, 3, 4)),
    by="name")

leafcutter.effects_and_sig %>%
  group_by(cluster, treatment) %>%
  mutate(ContainsGAGT = any(UpstreamOfDonor2BaseSeq=="GA")) %>%
  ungroup() %>%
  filter(p.adjust < 0.1) %>%
  distinct(cluster, treatment, .keep_all=T) %>%
  count(ContainsGAGT, treatment) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(x=treatment, y=n, fill=ContainsGAGT)) +
  geom_col(position="stack") +
  coord_flip() +
  theme(axis.text.y = element_text(size=6, colour = XaxisColors), legend.position='bottom') +
  labs(title="WD08 has many non-GAGT effects", y="Number significant clusters, FDR<10%")
```
Ok, so now I am sort of worried the WD08 effect might be indirect effects. There is much more AS but much of it is not at GAGT introns, and also there is tons of tons of DE.

The other two interesting molecules were WC05 and WB12. Those ones I still don't understand as well. Looking at the PCA again, WC05 is the larger deparature from the rsidiplam/branaplam axes than WB12. I'll keep a particular eye on WC05.

I'm going to look at the -3 and -4 positions now. So I'll just filter for the GAGT introns, and make a similar plot as the heatmap I plotted above for the -3 and -4 position ...

```{r, fig.width=7, fig.height=10}
leafcutter.effects_and_sig %>%
  filter(UpstreamOfDonor2BaseSeq == "GA") %>%
  mutate(NNGAGT = substr(seq, 1, 2)) %>%
  group_by(NNGAGT, treatment) %>%
  summarise(medLogef = median(logef)) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(y=treatment, x=NNGAGT, fill=medLogef)) +
  geom_raster() +
  scale_fill_viridis_c() +
  theme(axis.text.y = element_text(size=6, colour = XaxisColors))

```
Ok, so similar to previous Bhattacharya et al, and our dose titration series, we can see the -4 and -3 positions distinguish the branaplam-like molecules vs the risdiplam-like molecules. WC05 and WB12 don't look particularly special on this plot. If anything I think WC05 may look a bit more like risdiplam in this respect (I'm looking in particular at the AT column), consistent with its structural similarity with risdiplam. WC04 pops out as interesting, even though it didn't necessarily pop out from the PCA. I do notice in the above plots that WC04 had a lot of non GAGT diff splicing and a larger number of DE genes. Could also be some indirect effects.

Let's go back to the DE results. For now I'll plot volcano plots of some select molecules to see what the top genes are for some samples of interest: WC05, WB12, and some reference samples to represent risdiplam like effects (WB09, WB10) and branaplam like effects (WE02, WE04)

```{r, fig.width=10, fig.height=8}


diff.expression %>%
  filter(treatment %in% c("WC05", "WB12", "WE04", "WE02", "WB09","WB10")) %>%
  mutate(IsSig = FDR<0.1) %>%
  mutate(label = if_else(
    IsSig,
    symbol,
    NA_character_
  )) %>%
  mutate(treatment = recode(treatment,
    "WE02" = "WE02; branaplam-like",
    "WE04" = "WE04; branaplam-like",
    "WB09" = "WB09; risdiplam-like",
    "WB10" = "WB10; risdiplam-like",
  )) %>%
  ggplot(aes(x=logFC, y=-log10(PValue), color=IsSig)) +
  geom_point() +
  scale_color_manual(values=c(`TRUE`="red", `FALSE`="black")) +
  geom_text_repel(aes(label=label), size=3, max.overlaps=) +
  facet_wrap(~treatment, scales = "free") +
  labs("DE results, focusing on WB12 and WC05", color="FDR<10%")

```


Ok honing in on some genes, in some ways WC05 and WB12 are more risdiplam-like (DENND5A), and in some other ways they are more branaplam like (CRYL1), and in some ways they might be like niether (ZNF837)...


Let's plot these genes and a few others more closely...

```{r}
diff.expression %>%
  filter(symbol %in% c("HTT", "RCC1", "CRYL1", "ZNF837", "DENND5A", "FHOD3", "STAT1", "PITPNB")) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  mutate(Sig = case_when(
    PValue < 0.001 ~ "!",
    PValue < 0.05 ~ "*",
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x=treatment, y=logFC, fill=NewColor)) +
  geom_col(aes(alpha=alpha)) +
  geom_text(aes(label=Sig), y=Inf, vjust=1) +
  scale_fill_identity() +
  facet_wrap(~symbol) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=5), legend.position='none') +
  labs(title="Expression changes in genes of interest", y="log2FC", caption=str_wrap("P<0.05 = *\nP<0.001 = !\nColor and  fill transparency based on PC1 & PC2 to distinguish risdiplam/branaplam axes", 30))
```

There are a few treatments in this plot that also catch my eye in HTT and DENND5A... Let's look at those closer..

```{r}
diff.expression %>%
  filter(symbol %in% c("HTT", "DENND5A")) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33" | treatment %in% c("WC10", "WC11", "WC12"), 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  mutate(Sig = case_when(
    PValue < 0.001 ~ "!",
    PValue < 0.05 ~ "*",
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x=treatment, y=logFC, fill=NewColor)) +
  geom_col(aes(alpha=alpha)) +
  geom_text(aes(label=Sig), y=Inf, vjust=1) +
  scale_fill_identity() +
  facet_wrap(~symbol) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=5), legend.position='none') +
  labs(title="Expression changes in genes of interest", y="log2FC", caption=str_wrap("P<0.05 = *\nP<0.001 = !\nColor and  fill transparency based on PC1 & PC2 to distinguish risdiplam/branaplam axes. C10, C11, C12 reordered by reds", 30))
```



Let's repeat but just consider more systematically any DE gene found in C05 or B12 with FDR<20%

```{r, fig.height=8, fig.width=6}
diff.expression %>%
  filter(treatment %in% c("WC05", "WB12")) %>%
  filter( FDR<0.2) %>%
  distinct(EnsemblID) %>%
  left_join(diff.expression) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  mutate(Sig = case_when(
    PValue < 0.001 ~ "!",
    PValue < 0.05 ~ "*",
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x=treatment, y=logFC, fill=NewColor)) +
  geom_col(aes(alpha=alpha)) +
  geom_text(aes(label=Sig), y=Inf, vjust=1) +
  scale_fill_identity() +
  facet_wrap(~symbol) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=5), legend.position='none') +
  labs(title="Expression changes in genes of interest", y="log2FC", caption=str_wrap("P<0.05 = *\nP<0.001 = !\nColor and  fill transparency based on PC1 & PC2 to distinguish risdiplam/branaplam axes", 30))

# ggsave("../code/scratch/scratch.pdf", height=8, width=12)
```

Ok, based on this I'm not sure there are any astoundingly specific DE changes with WC05 or WB12... Sure in some ways it is unique becase it may be risdiplam like, and in some other ways it may be more branaplam like, and there are some changes in effect sizes, but no totally new DE discoveries.

I think it might be better to just report this to Jingxin before trying to make a list of things to verify by qPCR... That doesn't seem to interesting to me given these results.

Well, actually, let's at least do the analgous analysis but for splicing. And if that looks at all promising, then I can consider doing a more careful differential splicing or differential expression to directly compare the WC05 or WB12 treatments to more risdiplam like treatments or more branaplam like treatments to find what makes them unique.

**Still to do that...***

```{r}

```


Next thing I am curious about is following up on the WC04 effect at the -4 and -3 position. It seemed to have a preference for CGGA|GT and GGGA|GT that was not found in either risdiplam or branaplam like molecules. First let me verify this is really a believable effect and not just noise due to sampling a very low number of introns or something.

First let's see how many introns that median effect was based on and if it is significant...

For sake of not making the plot too big, let's just include a few hand picked dinucleotides based on previous plot.

```{r}
leafcutter.effects_and_sig %>%
  filter(UpstreamOfDonor2BaseSeq == "GA") %>%
  mutate(NNGAGT = substr(seq, 1, 2)) %>%
  filter(NNGAGT %in% c("AA","AT", "CA","CG", "GG", "TA", "TG", "TT")) %>%
  group_by(NNGAGT, treatment) %>%
  summarise(
    medLogef = median(logef),
    n=n(),
    MannWhitneyP=format.pval(wilcox.test(logef, alternative="greater")$p.value, 2)) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(y=treatment, x=NNGAGT, fill=medLogef)) +
  geom_raster() +
  geom_text(color='red', size=3, aes(label = paste(n, MannWhitneyP, sep='|'))) +
  scale_fill_viridis_c() +
  theme(axis.text.y = element_text(size=6, colour = XaxisColors))
```


Oh ok, actually those weird C04 effects are because that was the treatment where one of the replicates was a failed sample, that I have been excluding from some (but mistakenly not all) analyses.


Let's also plot the effects for that new bulge (guUaag) that I hypothesized in previous experiments is affected by these molecules (albeit to a lesser extent than gA|gu bulge).

```{r, fig.height=10, fig.width=12}
leafcutter.effects_and_sig %>%
  dplyr::select(treatment, intron, seq, logef) %>%
  pivot_wider(names_from = "treatment", values_from = "logef") %>%
  mutate(
    `1_ncag|guaagn` = str_detect(seq, "^\\wCAGGTAAG\\w"),
    `2_nagA|guaagn` = str_detect(seq, "^\\wAGAGTAAG\\w"),
    `3_nagG|guaagn` = str_detect(seq, "^\\wAGGGTAAG\\w"),
    `4_nagU|guaagn` = str_detect(seq, "^\\wAGTGTAAG\\w"),
    # `5.1_uagA|guaagn` = str_detect(seq, "^TAGAGTAAG\\w"),
    # `5.2_aagA|guaagn` = str_detect(seq, "^AAGAGTAAG\\w"),
    `5.3_cagA|guaagn` = str_detect(seq, "^CAGAGTAAG\\w"),
    `5.4_augA|guaagn` = str_detect(seq, "^ATGAGTAAG\\w"),
    `6_nnng|guUaag` = str_detect(seq, "^\\w\\w\\wGGTTAAG"),
    `7_nnng|guAaag` = str_detect(seq, "^\\w\\w\\wGGTAAAG"),
  ) %>%
  # filter_at(vars(matches("^\\d.+_.+$",perl=T)),any_vars(. == T)) %>%
  gather(key="Pattern", value="IsMatch", matches("^\\d.+",perl=T)) %>%
  # pull(Pattern) %>% unique()
  filter(IsMatch) %>%
  gather(key="treatment", value="logef", matches("^W.+")) %>%
  group_by(Pattern, treatment) %>%
  summarise(medLogef = median(logef, na.rm = T),
            p = format.pval(wilcox.test(logef, alternative="greater")$p.value, 2),
            n = sum(IsMatch)) %>%
  ungroup() %>%
  separate(Pattern, into = c("PlotOrder", "Pattern"), sep = "_", convert=T) %>%
  arrange(PlotOrder) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(y=treatment, x=Pattern, fill=medLogef)) +
  geom_raster() +
  geom_text(color='red', size=2, aes(label = paste(n, p, sep='|'))) +
  scale_fill_gradient2(name=str_wrap("Median dose-response effect", 10)) +
  Rotate_x_labels +
  theme(axis.text.y = element_text(size=6, colour = XaxisColors))

```


Ok, there really isn't anything to be said of that. We just lack power to say anything about the hypothesized effect on the guUaag bulge. We Again, we can't even really see the effects at the GT|GT introns which I previously saw at high concentrations and have been documented in other lit.

Ok, I think the last thing I want to try is to find the C05/B12-specific splicing changes by doing a direct contrast with branaplam-like and risdiplam-like molecules..

Let's replot the PCA, and highlight the samples I proposed picking for leafcutter contrasts. I'll do 3 contrasts for each sample of interest, labelled RisdiLike_High, RisdiLike_Med, and BranadLike_Med. I'll write out the leafcutter groups files for those contrasts, perform the contrast (in the snakemake), and then explore results in later code chunks:


```{r}

dir.create("../output/CustomLeafcutterGroupFiles")

Samples_of_interest <- c("WD08", "WC05", "WB12")
RisdiLike_Med_Samples <- c("WB06", "WB09", "WB10")
RisdiLike_High_Samples <- c("WD05", "WD07", "WB09")
BranaLike_Med_Samples <- c("WE04", "WE02", "WD10")


NewColors %>%
  dplyr::select(treatment, NewColor, alpha) %>%
  inner_join(pca.results.to.plot) %>%
ggplot(aes(x=PC1, y=PC2, color=NewColor)) +
  geom_point(size=3, aes(alpha=alpha)) +
  geom_text_repel(data = . %>%
            filter(treatment %in% c(Samples_of_interest, RisdiLike_Med_Samples)),
          aes(label=treatment)) +
  # geom_point(data = pca.results.to.plot_AdditionProjections) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Samples in RisdiLike_Med contrast labelled")

NewColors %>%
  dplyr::select(treatment, NewColor, alpha) %>%
  inner_join(pca.results.to.plot) %>%
ggplot(aes(x=PC1, y=PC2, color=NewColor)) +
  geom_point(size=3, aes(alpha=alpha)) +
  geom_text_repel(data = . %>%
            filter(treatment %in% c(Samples_of_interest, RisdiLike_High_Samples)),
          aes(label=treatment)) +
  # geom_point(data = pca.results.to.plot_AdditionProjections) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Samples in RisdiLike_High contrast labelled")

NewColors %>%
  dplyr::select(treatment, NewColor, alpha) %>%
  inner_join(pca.results.to.plot) %>%
ggplot(aes(x=PC1, y=PC2, color=NewColor)) +
  geom_point(size=3, aes(alpha=alpha)) +
  geom_text_repel(data = . %>%
            filter(treatment %in% c(Samples_of_interest, BranaLike_Med_Samples)),
          aes(label=treatment)) +
  # geom_point(data = pca.results.to.plot_AdditionProjections) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Samples in BranaLike_Med contrast labelled")

ControlSamples <- data.frame(BranaLike_Med_Samples, RisdiLike_High_Samples, RisdiLike_Med_Samples) %>%
  gather(key="ContrastName", value = "treatment") %>%
  mutate(Group = "Control") %>%
  left_join(FullMetadata) %>%
  dplyr::select(old.sample.name, Group, ContrastName)

for (treatment in Samples_of_interest){
  data.frame(treatment=treatment, ContrastName = c("BranaLike_Med_Samples", "RisdiLike_High_Samples", "RisdiLike_Med_Samples")) %>%
    left_join(FullMetadata) %>%
    dplyr::select(old.sample.name, ContrastName) %>%
    mutate(Group = "Treatment") %>%
    bind_rows(ControlSamples) %>%
    arrange(ContrastName, Group, old.sample.name) %>%
    group_by(ContrastName) %>%
    group_walk(~ write_tsv(.x, paste0("../output/CustomLeafcutterGroupFiles/Contrast_", treatment, "_vs_",.y$ContrastName, "groups.tsv"), col_names = F))
}



```

And then I added rules to the snakemake pipeline to run these contrasts:

```{bash, eval=F}
snakemake GatherCustomContrasts
```


Now let's explore the results..

```{r}
leafcutter.effects.CustomContrasts <- list.files(path="../code/SplicingAnalysis/Exp52_CustomContrasts", pattern="^.+_effect_sizes.txt", full.names=T) %>%
  setNames(str_replace(., "../code/SplicingAnalysis/Exp52_CustomContrasts/(.+?)_effect_sizes.txt", "\\1")) %>%
  lapply(fread) %>%
  bind_rows(.id="Contrast") %>%
  mutate(name = str_replace(intron, "(.+?):(.+?):(.+?):clu.+([+-])", "\\1_\\2_\\3_\\4")) %>%
  mutate(cluster = str_replace(intron, "(.+?):.+?:.+?:(clu.+[+-])", "\\1:\\2"))

leafcutter.sig.CustomContrasts <-list.files(path="../code/SplicingAnalysis/Exp52_CustomContrasts", pattern="^.+_cluster_significance.txt", full.names=T) %>%
  setNames(str_replace(., "../code/SplicingAnalysis/Exp52_CustomContrasts/(.+?)_cluster_significance.txt", "\\1")) %>%
  lapply(fread) %>%
  bind_rows(.id="Contrast")

leafcutter.effects.Merged.CustomContrasts <- leafcutter.effects.CustomContrasts %>%
  inner_join(leafcutter.sig.CustomContrasts, by=c('cluster', 'Contrast')) %>%
  left_join(Introns.annotations, by="name") %>%
  mutate(
    TestTreatment = str_replace(Contrast,"^Contrast_(.+?)_vs_(.+?)_Samplesgroups$", "\\1"),
    ControlTreatment = str_replace(Contrast,"^Contrast_(.+?)_vs_(.+?)_Samplesgroups$", "\\2"),
    UpstreamOfDonor2BaseSeq = substr(seq, 3, 4)
  )

leafcutter.effects.Merged.CustomContrasts$TestTreatment %>% unique()

leafcutter.effects.Merged.CustomContrasts %>%
  group_by(cluster, TestTreatment, ControlTreatment) %>%
  mutate(ContainsGAGT = any(UpstreamOfDonor2BaseSeq=="GA")) %>%
  ungroup() %>%
  filter(p.adjust < 0.1) %>%
  distinct(cluster, TestTreatment, ControlTreatment, .keep_all=T) %>%
  count(ContainsGAGT, TestTreatment, ControlTreatment) %>%
  ggplot(aes(x=ControlTreatment, y=n, fill=ContainsGAGT)) +
  geom_col(position="stack") +
  facet_wrap(~TestTreatment, scales="free") +
  theme(legend.position='bottom') +
  Rotate_x_labels +
  labs(title=str_wrap("contrasts vs brana-like and risdi-like treatments", 40), y="Number significant clusters, FDR<10%")

```

Ok, so there are fair number of significant differences. That was totally expected. but of course I am particulary interseted in the GAGT introns that are *more activated* in C05/D08/B12

```{r}
leafcutter.effects.Merged.CustomContrasts %>%
  group_by(cluster, TestTreatment, ControlTreatment) %>%
  mutate(ContainsGAGT = any(UpstreamOfDonor2BaseSeq=="GA")) %>%
  ungroup() %>%
  filter(p.adjust < 0.1) %>%
  filter(ContainsGAGT) %>%
  mutate(sign = sign(logef)) %>%
  count(sign, TestTreatment, ControlTreatment) %>%
  ggplot(aes(x=ControlTreatment, y=n, fill=as.factor(sign))) +
  geom_col(position="stack") +
  facet_wrap(~TestTreatment, scales="free") +
  theme(legend.position='bottom') +
  Rotate_x_labels +
  labs(title=str_wrap("contrasts vs brana-like and risdi-like treatments", 40), y="Number significant GAGT introns, FDR<10%")
```

Ok, about the same number of upregulated vs downregulated introns in each contrast. What I am really after I should say are the introns that are upregulated in all contrasts. So I'll filter for GA|GT introns that are FDR<10% significant in at least one contrast, and nominally significant in all contrasts (P<0.05), and with a effect size direction favoring upregulation in the treatment of interest.

```{r, fig.height=5, fig.width=10}
IntronsActivatedByTreatmentsOfInterest <- leafcutter.effects.Merged.CustomContrasts %>%
  filter(UpstreamOfDonor2BaseSeq == "GA") %>%
  group_by(intron, TestTreatment) %>%
  filter(any(p.adjust < 0.1)) %>%
  filter(all(p < 0.05)) %>%
  filter(all(logef > 0)) %>%
  filter(all(deltapsi > 0.05)) %>%
  ungroup()

IntronsActivatedByTreatmentsOfInterest %>%
  dplyr::select(intron, Treatment, Control, TestTreatment, ControlTreatment ) %>%
  group_by(TestTreatment, intron) %>%
  mutate(MeanTreatment = mean(Treatment)) %>%
  distinct(intron, TestTreatment, ControlTreatment, .keep_all=T) %>%
  dplyr::select(-Treatment) %>%
  pivot_wider(names_from = "ControlTreatment", values_from="Control") %>%
  gather(key="Group", value="IntronicPSI", -c("intron", "TestTreatment")) %>%
  mutate(Group= relevel(factor(Group), ref="MeanTreatment")) %>%
  mutate(label = paste(TestTreatment, intron, sep='\n')) %>%
  ggplot(aes(x=Group, y=IntronicPSI, fill=Group)) +
  geom_col() +
  facet_wrap(~label, scales="free") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text.x = element_text(size = 5)) +
  labs(title="GA-GT introns most specific to samples of interest", caption=str_wrap("DeltaPSI>5% in all contrasts, FDR<10% in at least one contrast, P<0.05 in all contrasts", 40))

# ggsave("../code/scratch/scratch.pdf", height=6, width=12)
```

Ok these are the kinds of introns I'm looking for, that Jingxin can consider validating as specific to these three molecules of interest. Let's look the original data for each sample for these sets of introns... And then double check in the genome browser.

```{r, fig.height=7, fig.width=8}
IntronsActivatedByTreatmentsOfInterest.vec <- IntronsActivatedByTreatmentsOfInterest %>% pull(intron) %>% unique()

AllSamplesInContrasts <- c(Samples_of_interest, RisdiLike_Med_Samples, RisdiLike_High_Samples, BranaLike_Med_Samples)


PSI %>%
  filter(junc %in% IntronsActivatedByTreatmentsOfInterest.vec) %>%
  dplyr::select(junc, contains("NewMolecule")) %>%
  gather(key = "leafcutter.name", value="IntronicPSI", -junc) %>%
  left_join(FullMetadata, by="leafcutter.name") %>%
  inner_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33" | treatment %in% c("WC10", "WC11", "WC12"), 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  filter(treatment %in% AllSamplesInContrasts) %>%
  # sample_n_of(5, junc) %>%
  ggplot(aes(x=treatment, y=IntronicPSI, color=NewColor)) +
  geom_text(aes(alpha=alpha, label=treatment), size=3) +
  facet_wrap(~junc, scales="free") +
  scale_color_identity() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text.x = element_text(size = 5),
        legend.position='none') +
  labs(title="Candidate treatment-of-interest-specific effects", x="Treatments, roughly in order", y="Intronic PSI")
```

Ok, I guess that seems reasonable based on the previous plots. I don't think there is a bug in my code.

Ok, let's include all samples in these plots now, to see to what extent these candidates are just specific to the molecule of interest.

```{r}
PSI %>%
  filter(junc %in% IntronsActivatedByTreatmentsOfInterest.vec) %>%
  dplyr::select(junc, contains("NewMolecule")) %>%
  gather(key = "leafcutter.name", value="IntronicPSI", -junc) %>%
  left_join(FullMetadata, by="leafcutter.name") %>%
  inner_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33" | treatment %in% c("WC10", "WC11", "WC12"), 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  # sample_n_of(5, junc) %>%
  ggplot(aes(x=treatment, y=IntronicPSI, color=NewColor)) +
  geom_text(aes(alpha=alpha, label=treatment), size=3) +
  facet_wrap(~junc, scales="free") +
  scale_color_identity() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text.x = element_text(size = 5),
        legend.position='none') +
    labs(title="Candidate treatment-of-interest-specific effects", x="Treatments, roughly in order", y="Intronic PSI")

```

Ok, there are a few D08 specific effects, but perhaps that isn't surprising given previous plots that show so many effects in this particular treatment.

In any case, as promised, I'll write out some of the most promising things, including FHOD3, as junctions worthy of validation... Maybe the better approach for identifying the worthy candidate list this is just to forget these specific contrasts, and look at the DE plots I made, and intersect with potential GA|GT intron drivers...



```{r}

GenesToHighlight <- c("HSPA1L", "CRYL1", "FHOD3", "ZFYVE26")

diff.expression %>%
  filter(symbol %in% GenesToHighlight) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  mutate(Sig = case_when(
    PValue < 0.001 ~ "!",
    PValue < 0.05 ~ "*",
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x=treatment, y=logFC, fill=NewColor)) +
  geom_col(aes(alpha=alpha)) +
  geom_text(aes(label=Sig), y=Inf, vjust=1) +
  scale_fill_identity() +
  facet_wrap(~symbol) +
  Rotate_x_labels +
  theme(axis.text.x = element_text(size=5), legend.position='none') +
  labs(title="Expression changes in genes of interest", y="log2FC", caption=str_wrap("P<0.05 = *\nP<0.001 = !\nColor and  fill transparency based on PC1 & PC2 to distinguish risdiplam/branaplam axes", 30))
```

These 4 genes are examples of how these red treatments may sometimes be more branaplam like on some genes, and more risdiplam like on others. Now let's try and find the GA-GT introns drivers for these genes...

For convenience let's look from the data table I wrote out from estimating EC50s from the dose response experiment

```{r}

GA.GT.introns_FromDoseResponseExperiment <- read_tsv("../output/EC50Estimtes.FromPSI.txt.gz")

GA.GT.introns_FromDoseResponseExperiment %>%
  separate_rows(gene_names, sep=",") %>%
  filter(gene_names %in% GenesToHighlight)
  

```

Ok actually that didn't find much... And according to that previous data, CRYL1 GA-GT intron as actually more risdiplam-specific... weird. Ok, let's just go back to trying to find the GA-GT intron from the leafcutter results...

```{r}
leafcutter.effects_and_sig %>%
  filter(UpstreamOfDonor2BaseSeq=="GA") %>%
  separate_rows(gene_names, sep=",") %>%
  filter(gene_names %in% GenesToHighlight) %>%
  distinct(intron, gene_names) %>%
  kable()

leafcutter.effects_and_sig %>%
  filter(UpstreamOfDonor2BaseSeq=="GA") %>%
  separate_rows(gene_names, sep=",") %>%
  filter(gene_names %in% GenesToHighlight) %>%
  # distinct(intron, .keep_all=T) %>%
  dplyr::select(intron, treatment_PSI, DMSO_PSI, treatment, gene_names) %>%
  group_by(intron) %>%
  mutate(DMSO_PSI = mean(DMSO_PSI)) %>%
  ungroup() %>%
  pivot_wider(names_from="treatment", values_from="treatment_PSI", names_glue="{treatment}_PSI") %>%
  dplyr::select(intron, gene_names, everything()) %>%
  gather(key="treatment", value="IntronPSI", -c("gene_names", "intron")) %>%
  mutate(treatment = str_replace(treatment, "^(.+?)_PSI", "\\1")) %>%
  mutate(label=paste(intron, gene_names, sep='\n')) %>%
  left_join(NewColors) %>%
  mutate(PC2 = if_else(NewColor == "#e34a33", 0, PC2)) %>%
  mutate(treatment = fct_reorder(treatment, PC2 * alpha)) %>%
  ggplot(aes(x=treatment, y=IntronPSI, fill=NewColor)) +
  geom_col(aes(alpha=alpha)) +
  scale_fill_identity() +
  facet_wrap(~label, scales="free") +
  Rotate_x_labels +
  theme(legend.position='none', axis.text.x = element_text(size=5))
```

Ok, so I recovered the same CRYL1 intron, and also the FHOD3 intron. Still couldn't find the intron for the other two genes... I think I am satisfied enough just to say these two introns might be good to validate molecule-specific effects, and also probably any number of the risdiplam-specific or branaplam-specific introns that I previously shared from the dose-response data.

Also, looking at the genome browser of these two introns listed above it becomes obvious what the exon coordinates are if you were to design qPCR probes for validation... FHOD3 is particularly interesting... Much like I previously observed in HTT, the actual intron with the GA-GT 5'ss seems to be mostly retained in the treatment - it is the upstream cryptic intron that more obviously gets excised in the treatment. This suggests that the treatment may aid in U1 binding and activation of exon definition without necessarily activating the 5'ss bound to the molecule. This may also somewhat explain my previous observation that these molecules induce much cryptic 3'ss that aren't in a leafcutter cluster with a cryptic 5'ss.

Here is an screenshot of what I mean.

[![FHOD3](assets/FHOD3_Example)]

### PCA on subset of scaffolds

Jingxin shared a powerpoint in which he color coded samples based on their scaffolds... Using these and some of my custom observations, I wrote factors for each treatment. I'm going to save that data replot PCA a few different ways for exploration.

```{r}

TreatmentFactors <- read_csv("../code/DataNotToCommit/CompoundsAndManualAnnotationFactors.csv") %>%
  mutate(treatment = paste0("W", Well))

```

First let me replot the PCA the way I have been plotting it previously with colors consistent with previous plots

```{r}
pca.results.to.plot %>%
  left_join(
    NewColors %>%
      dplyr::select(NewColor, alpha, treatment)) %>%
ggplot(aes(x=PC1, y=PC2, color=NewColor, alpha=alpha)) +
  geom_text(size=2, aes(label=treatment)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Colored consistent in other plots")


pca.results.to.plot %>%
  left_join(TreatmentFactors) %>%
ggplot(aes(x=PC1, y=PC2, color=Factor)) +
  geom_text(size=2, aes(label=treatment)) +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Colored by structural groupings from Jingxins")


pca.results.to.plot %>%
  left_join(TreatmentFactors) %>%
ggplot(aes(x=PC3, y=PC4, color=Factor)) +
  geom_text(size=2, aes(label=treatment)) +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nAll samples", caption="Colored by structural groupings from Jingxins")
```

Ok, one thing I want to try is to repeat the PCA but only exclude the branaplam-based molecules.. I think I will project the dose-response experiment data to aid in interpretation.


```{r}
NonBranaplamBasedMolecules <- TreatmentFactors %>%
  filter(!Factor=="G") %>% pull(treatment)

SamplesToInclude <- 
  FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  filter(treatment %in% c("DMSO", NonBranaplamBasedMolecules)) %>%
  pull(leafcutter.name)
  

pca.results <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata %>%
              dplyr::select(leafcutter.name, treatment, SampleName)) %>%
  left_join(TreatmentFactors, by="treatment") %>%
  mutate(ToHighlight = if_else(
    treatment %in% c("WB12", "WC05", "WD08"),
    1,
    0.5
    ))

SamplesToProject <- FullMetadata %>%
  filter(Experiment %in% c("Dose response titration")) %>%
  pull(leafcutter.name)

pca.results.to.plot_AdditionProjections <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToProject)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% predict(pca.results, .) %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

P <- ggplot(pca.results.to.plot, aes(x=PC1, y=PC2)) +
  geom_text(size=3, aes(label=treatment, color=Factor, alpha=ToHighlight)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")
P
P +
  new_scale_colour() +
  geom_point(data = pca.results.to.plot_AdditionProjections,
             aes(color=color)) +
  scale_color_identity()


P <- ggplot(pca.results.to.plot, aes(x=PC3, y=PC4)) +
  geom_text(size=3, aes(label=treatment, color=Factor, alpha=ToHighlight)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")
P
P +
  new_scale_colour() +
  geom_point(data = pca.results.to.plot_AdditionProjections,
             aes(color=color)) +
  scale_color_identity()

ggplot(pca.results.to.plot, aes(x=PC5, y=PC6)) +
  geom_text(size=3, aes(label=treatment, color=Factor, alpha=ToHighlight)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  new_scale_colour() +
  geom_point(data = pca.results.to.plot_AdditionProjections,
             aes(color=color)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")

pca.results.to.plot %>%
  inner_join(
    counts.plot.dat %>%
      dplyr::select(SampleName, lib.size)
    ) %>%
ggplot(aes(x=PC5, y=PC6, color=lib.size)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  scale_color_viridis_c() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")
```

And let's do the same for branaplam based molecules

```{r}
BranaplamBasedMolecules <- TreatmentFactors %>%
  filter(Factor=="G") %>% pull(treatment)

SamplesToInclude <- 
  FullMetadata %>%
  filter(Experiment %in% c("Single high dose LCL")) %>%
  filter(!old.sample.name == "NewMolecule.C04-2") %>%
  filter(treatment %in% c("DMSO", BranaplamBasedMolecules)) %>%
  pull(leafcutter.name)
  

pca.results <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToInclude)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% prcomp(scale=T)


pca.results.to.plot <- pca.results$x %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata %>%
              dplyr::select(leafcutter.name, treatment, SampleName)) %>%
  left_join(TreatmentFactors, by="treatment") %>%
  mutate(ToHighlight = if_else(
    treatment %in% c("WB12", "WC05", "WD08"),
    1,
    0.5
    ))

SamplesToProject <- FullMetadata %>%
  filter(Experiment %in% c("Dose response titration")) %>%
  pull(leafcutter.name)

pca.results.to.plot_AdditionProjections <- PSI.IntronsToInclude %>%
  dplyr::select(junc, all_of(SamplesToProject)) %>%
  column_to_rownames("junc") %>%
  scale() %>% t() %>% predict(pca.results, .) %>%
  as.data.frame() %>%
  rownames_to_column("leafcutter.name") %>%
  dplyr::select(leafcutter.name, PC1:PC6) %>%
  left_join(FullMetadata, by="leafcutter.name")

P <- ggplot(pca.results.to.plot, aes(x=PC1, y=PC2)) +
  geom_text(size=3, aes(label=treatment, color=Factor, alpha=ToHighlight)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")
P
P +
  new_scale_colour() +
  geom_point(data = pca.results.to.plot_AdditionProjections,
             aes(color=color)) +
  scale_color_identity()


P <- ggplot(pca.results.to.plot, aes(x=PC3, y=PC4)) +
  geom_text(size=3, aes(label=treatment, color=Factor, alpha=ToHighlight)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")
P
P +
  new_scale_colour() +
  geom_point(data = pca.results.to.plot_AdditionProjections,
             aes(color=color)) +
  scale_color_identity()

ggplot(pca.results.to.plot, aes(x=PC5, y=PC6)) +
  geom_text(size=3, aes(label=treatment, color=Factor, alpha=ToHighlight)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  new_scale_colour() +
  geom_point(data = pca.results.to.plot_AdditionProjections,
             aes(color=color)) +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")

pca.results.to.plot %>%
  inner_join(
    counts.plot.dat %>%
      dplyr::select(SampleName, lib.size)
    ) %>%
ggplot(aes(x=PC5, y=PC6, color=lib.size)) +
  geom_text(size=3, aes(label=treatment)) +
  scale_alpha_continuous(range=c(0.5,1)) +
  scale_color_viridis_c() +
  theme_bw() +
  labs(title = "PCA using 632 upregulated GAGT introns\nNon branaplam skeletons", caption="Colored by structural groupings from Jingxins")


```

I'm re-questioning whether C05 and B12 are special in any worthwhile way


