---
title: "Plot PCA with structures"
output: html_document
date: "2023-02-23"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

library(tidyverse)
library(RColorBrewer)
library(data.table)
library(ggimage)



# define some useful funcs
sample_n_of <- function(data, size, ...) {
  dots <- quos(...)
  
  group_ids <- data %>% 
    group_by(!!! dots) %>% 
    group_indices()
  
  sampled_groups <- sample(unique(group_ids), size)
  
  data %>% 
    filter(group_ids %in% sampled_groups)
}


# Set theme
theme_set(
  theme_classic() +
  theme(text=element_text(size=16,  family="Helvetica")))

# I use layer a lot, to rotate long x-axis labels
Rotate_x_labels <- theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

#test plot
ggplot(mtcars, aes(x=mpg, y=cyl)) +
  geom_point()


```

## Intro

I previously performed a PCA to visualize the 52 compounds, relative to risdiplam/branaplam/C2C5 titrations. Here I want to plot that data with chemical structures for points.

```{r}
FullMetadata <- read_tsv("../output/20230223_FullMetadata.tsv")

PSI.IntronsToInclude <- read_tsv("../output/20230223_IntronsToInclude.tsv.gz")

PCA.dat <- read_tsv("../output/20230227_PCA_Projections.tsv.gz")

MoleculeIdentities <- read_csv("../code/DataNotToCommit/CompoundsAndConcentrations_ZT.csv")


Images <- data.frame( Images = Sys.glob("../code/DataNotToCommit/4x/*.png") ) %>%
  mutate(CompoundName = str_replace(Images, "../code/DataNotToCommit/4x/(.+?).png", "\\1")) %>%
  left_join(MoleculeIdentities)

data.frame( Images = Sys.glob("../code/DataNotToCommit/4x/*.png") ) %>%
  mutate(CompoundName = str_replace(Images, "../code/DataNotToCommit/4x/(.+?).png", "\\1")) %>%
  full_join(MoleculeIdentities) %>%
  filter(is.na(treatment) | is.na(Images))

```

There will be a few too many points to make sense of the plot imo. For now, (hackish solution) I will just average PC values across biological replicates and plot just one point per treatment.

```{r}
TreatmentColorsForLabels.FibroblastColorsAdded <-
FullMetadata %>%
  group_by(treatment) %>%
  filter(dose.nM == max(dose.nM) | treatment == "DMSO" | cell.type=="Fibroblast") %>%
  ungroup() %>%
  distinct(treatment, .keep_all=T) %>%
  arrange(dose.nM) %>%
  mutate(vjust=row_number()*1.2)

TreatmentColorsLabels.Layer.FibroblastColorsAdded <- geom_text(
    data = TreatmentColorsForLabels.FibroblastColorsAdded, 
    aes(label=treatment, color=color, vjust=vjust),
    y=Inf, x=Inf, hjust=1.05
  )

  


pca.results.to.plot <- PCA.dat %>%
  group_by(treatment, dose.nM, color, Experiment) %>%
  summarise(PC1 = mean(PC1), PC2=mean(PC2), PC3=mean(PC3), PC4=mean(PC4), PC5=mean(PC5)) %>%
  ungroup() %>%
  left_join(Images %>%
              dplyr::select(-dose.nM), by="treatment")


PC1.PC2 <- ggplot(pca.results.to.plot, aes(x=PC1, y=PC2, shape=Experiment)) +
  geom_image(aes(image=Images), size=0.14) +
  # geom_text(aes(label=treatment)) +
  geom_point(data = . %>%
               filter(Experiment %in% c("Dose response titration", "Single high dose fibroblast")), aes(color=color)) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  xlim(c(-20, 42)) +
  theme_bw() +
  labs(title = "PCA using 475 GAGT introns\nAll samples", caption="Introns based on being significant in any contrast") +
  theme(legend.position='none')
PC1.PC2


PC3.PC4 <- ggplot(pca.results.to.plot, aes(x=PC3, y=PC4, shape=Experiment)) +
  geom_image(aes(image=Images), size=0.15) +
  geom_point(data = . %>%
               filter(Experiment %in% c("Dose response titration", "Single high dose fibroblast")), aes(color=color)) +
  TreatmentColorsLabels.Layer.FibroblastColorsAdded +
  scale_color_identity() +
  theme_bw() +
  labs(title = "PCA using 632 GAGT introns\nAll samples", caption="Introns based on being significant in any contrast")


```

```{r, eval=F}
ggsave("../code/scratch/PC1.PC2.png",PC1.PC2, height=8, width=12)

```


Ok, the only thing that is annoying about this plot is there are too many overlapping sticks. maybe a more useful way to plot this is as a heatmap with each skeleton in their own box, and fill the box by the color based on PC1/PC2...

```{r}
NewColors <- PCA.dat %>%
  filter(Experiment == "Single high dose LCL") %>%
  # group_by(treatment) %>%
  # summarise(PC1 = mean(PC1), PC2=mean(PC2)) %>%
  mutate(NewColor = case_when(
    PC2 > 0 ~ "#08519c",
    PC2 < 0 ~ "#006d2c",
  ))

MinPC1 <- min(NewColors$PC1)
MaxPC1 <- max(NewColors$PC1)

NewColors <- NewColors %>%
  mutate(alpha = (PC1 - MinPC1)/(MaxPC1 - MinPC1)) %>%
  mutate(NewColor = if_else(
    treatment %in% c("WD08", "WC05", "WB12"),
    "#e34a33",
    NewColor
  ))

PCA.TwoColorsAndFill <- NewColors %>%
  mutate(treatment=str_remove_all(treatment, "W")) %>%
  ggplot(aes(x=PC1, y=PC2, color=PC2, alpha=PC1)) +
  geom_text(aes(label=treatment),size=3) +
  scale_colour_gradient(low="#e34a33", high="#08519c") +
  # scale_color_gradient2(low="#006d2c", mid="#e34a33", high="#08519c", midpoint=-5) +
  scale_alpha_continuous(range=c(0, 1))
PCA.TwoColorsAndFill

Heatmap.Plate <- NewColors %>%
  dplyr::select(-dose.nM) %>%
  left_join(MoleculeIdentities) %>%
  mutate(Row_Column = str_replace(leafcutter.name, "NewMolecule.([A-E])(\\d+).[1-6]$", "\\1_\\2")) %>%
  replace_na(list(CompoundName="DMSO")) %>%
  separate(Row_Column, into=c("Row", "Column"), sep="_", convert=T) %>%
  group_by(Row, Column, treatment) %>%
  mutate(rep = row_number()) %>%
  ungroup() %>%
  left_join(Images) %>%
  ggplot(aes(x=rep, y=1)) +
  geom_col(aes(fill=PC2, alpha=PC1), position = "fill", width=1) + 
  geom_text(data = . %>%
              filter(rep==1),
            aes(label=CompoundName), x=1, y=-Inf, hjust=0.2, vjust=-1, size=2.5) +
  geom_image(data = . %>%
               filter(rep==1),
             aes(image=Images), size=0.9, y=0.7, x=1, nudge_x=0.5) +
  scale_fill_gradient(low="#e34a33", high="#08519c") +
  # scale_fill_gradient2(low="#006d2c", mid="#e34a33", high="#08519c", midpoint=-5) +
  scale_alpha_continuous(range=c(0, 1)) +
  facet_grid(Row~Column) +
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_continuous(expand = c(0, 0)) +
  theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        aspect.ratio = 1)
Heatmap.Plate

```

```{r, eval=F}
ggsave("../code/scratch/PlateLayout_PCA_Heatmap.png",Heatmap.Plate, height=6, width=12)
ggsave("../code/scratch/PlateLayout_PCA_TwoColors.png",PCA.TwoColorsAndFill, height=6, width=7)

```


